Analysis of cloud enhancement events in a 30-year record of global solar irradiance at Thessaloniki, Greece

### 2024-08-31 article/article.Rmd
Commit 2024-08-31 13:12

author:
  - name:        Athanasios N. Natsis
    email:       natsisphysicist@gmail.com
    affiliation: LAP
    correspondingauthor: true
    # footnote: 1
  - name:        Alkiviadis Bais
    email:       abais@auth.gr
    affiliation: LAP
  - name:        Charikleia Meleti
    email:       meleti@auth.gr
    affiliation: LAP
    # footnote: 1
address:
  - code:         LAP
    organization: Laboratory of Atmospheric Physics, Physics Department, Aristotle University of Thessaloniki
    # addressline:  1 main street
    city:         Thessaloniki
    postcode:     54124
    country:      Greece
  # - code:         Another University
  #   organization: Department
  #   addressline:  A street 29
  #   postcode:     2054 NX
  #   city:         Manchester,
  #   state:        State
  #   country:      The Netherlands
# footnote:
#   - code: 1
#     text: "This is the first author footnote."
#   - code: 2
#     text: "Another author footnote."
abstract: |
  In this study we investigate the characteristics of global horizontal irradiance
  enhancement events induced by clouds over Thessaloniki for the period 1993 -- 2023
  using data recorded every one minute. We identified the cloud enhancement (CE)
  events by creating an appropriate clear-sky irradiance reference with the use of a
  radiative transfer model and aerosol optical depth data from a collocated Cimel sun
  photometer and a Brewer spectrophotometer. We found a trend in CE events of
  \(45.6\pm 21.9\,\text{cases}/\text{year}\), and a trend in the CE events
  irradiation of \(116.9\pm 67.8\,\text{kJ}/\text{year}\). The peak of the CE events
  was observed during May and June. The analysis of the total duration of CE events
  showed that durations longer than 5 minutes are very rare, with exceptions lasting
  over an hour and up to 140 minutes. Finally, we have detected enhancements above
  the total solar irradiance at the top of the atmosphere of up to
  \(400\,\text{W}/\text{m}^{2}\), with the \(75\,\%\) of the cases below
  \(200\,\text{W}/\text{m}^{2}\). The most active period of these extreme events is
  spring -- early summer with a secondary peak in autumn.

@@

We studied the enhancement of the global horizontal irradiance induced by clouds over
Thessaloniki for the period 1993 -- 2023 using data recorded every one minute.  We
identified the cloud enhancement events (CE) by creating an appropriate clear sky
irradiance reference with the use of a radiative transfer model and aerosol optical
depth data from a collocated Cimel sun photometer and a Brewer spectrophotometer. We
found a trend in CE events of
$`r round(yearlyCE[var == "GLB_diff.N", slope], 0)`\pm
`r round(2 * yearlyCE[var == "GLB_diff.N", slope.sd],0)`\,\text{cases}/\text{year}$,
and a trend in the CE events irradiation of 
$`r round(yearlyCE[var == "GLB_diff.sum", slope], 1)`\pm
`r round(2 * yearlyCE[var == "GLB_diff.sum", slope.sd], 1)`\,\text{kJ}/\text{year}$.
The peak of the CE events was observed during May and June.  An analysis of the total
duration of CE events, showed that durations longer than 5 minutes are very rare,
with exceptions lasting over an hour and up to 140 minutes.  Finally, we have
detected enhancements above the total solar irradiance at the top of the atmosphere
for the same solar zenith angle of up to
$400\,\text{W}/\text{m}^{2}$, with the $75\,\%$ of the cases below
$200\,\text{W}/\text{m}^{2}$. The most active season of these extreme events is
spring until early summer and autumn.

@@

<!-- Radiation data quality check -->
To further improve the quality of the irradiance data, a manual screening was
performed, in order to remove inconsistent and erroneous recordings that can occur
stochastically or systematically during the long operation of the instrument.  The
manual screening was aided by a radiation data quality assurance procedure, adjusted
for the site, which was based on the methods of Long and Shi\ [@Long2006;
@Long2008a].  Thus, problematic recordings have been excluded from further
processing.  Furthermore, due to the significant measurement uncertainty in GHI when
the Sun is near the horizon, and due to some systematic obstructions by nearby
buildings, we have excluded all measurements with solar zenith angle (SZA) greater
than $`r 90 - BIO_ELEVA`^\circ$.  Finally, images from a sky camera have been used in
the manual inspection of the CE identification. The sky camera operates since 2012
and stores images in 5 min time steps.
<!--
Although it is impossible to detect all false data, the large number of
available data, and the aggregation scheme we used, ensures the quality of the
radiation measurements used in this study.
-->
<!-- Data selection -->
<!--
To preserve an unbiased representation of the data we applied a constraint, similar
the one used by @CastillejoCuberos2020. For each valid hour of day, there must exist
at least 45 minutes of valid measurements, including nighttime, when near sunrise and
sunset. Days with less than 5 valid hours were rejected completely.
-->
<!--
# CastillejoCuberos2020 There must be at least 5 h of valid data (Tests 1–7),
otherwise all data for this day is flagged as invalid. There must be at least This
work valid data 45 min of valid data (Tests 1–7) for an hour to be flagged as valid,
otherwise all minute data for this hour is flagged as invalid.
-->

@@

<!-- Aerosol change -->
The clear-sky reference values discussed above are based on the climatological AOD
and WC; hence they cannot describe accurately the long-term variation of
$\text{GHI}_\text{ref}$ due to long-term changes in the two atmospheric constituents,
mainly AOD.  As reported by @Natsis2023, there is a long-term brightening effect in
the GHI data of Thessaloniki for the period 1993 -- 2023, which for clear-sky data
was attributed to long-term changes in aerosol effects.  Therefore, an adjustment of
the $\text{GHI}_\text{ref}$ during the period of study was made using simulations
with the RTM based on the long-term variations of the AOD. As AERONET data start only
in 2003, we used for the period 1993 -- 2005 estimates of changes in AOD at
$340\,\text{nm}$ derived from a collocated Brewer spectrophotometer [@Kazadzis2007]
to calculate the change in $\text{GHI}_\text{ref}$ due to aerosols during this period.

@@

```{r CS-change, echo=FALSE, fig.cap="Simulated long-term change in clear sky irradiance relative to the climatological values due to changes in AOD in Thessaloniki for the period 1993 -- 2023."}
knitr::include_graphics("../images/P-CS-change-1.pdf")
```

@@

Figure\nobreakspace{}\ref{fig:P-energy} shows the time series of the yearly number of
CE cases (each with duration of one minute), the yearly mean OI and the yearly
excess irradiation for the period 1993 -- 2023, together with corresponding linear
trends. All three quantities show increasing trends, most pronounced for the
frequency of occurrence 
($`r plus_sign(round(yearlyCE[var == "GLB_diff.N", slope], 0))`\pm `r round(2 * yearlyCE[var == "GLB_diff.N", slope.sd], 0)`\,\text{cases}/\text{year}$)
and the excess irradiation 
($`r plus_sign(round(yearlyCE[var == "GLB_diff.sum", slope], 1))` \pm `r round(2 * yearlyCE[var == "GLB_diff.sum", slope.sd], 1)`\,\text{kJ}/\text{year}$),
which are also statistically significant.
In contrast the trend of the yearly mean OI is negligible 
($`r plus_sign(round(yearlyCE[var == "GLB_diff.mean", slope], 2))`\pm `r round(2 * yearlyCE[var == "GLB_diff.mean", slope.sd], 1)`\,\text{W}/\text{m}^2/\text{year}$)
and of no statistical significance.
The average OI for the entire period is
$`r plus_sign(round(mean(ST_E_yearly$GLB_diff.mean), 1))`\pm `r round(sd(ST_E_yearly$GLB_diff.mean),1)`\,\text{W}/\text{m}^2$.
The interannual variability of the data about the trend lines is quite large.
Furthermore, the spread tends to increase with time (at least for the quantities of
panels b and c), suggesting a significant variability in cloud patterns over the
area, possibly associated to changes in climate.

@@

\begin{figure}% [h!]
        {\centering 
            \subfloat[\label{fig:P-energy-mean}]
                {\includegraphics[width=\linewidth]{../images/P-energy-3} }\\
            \subfloat[\label{fig:P-energy-N}]
                {\includegraphics[width=\linewidth]{../images/P-energy-2} }\\
            \subfloat[\label{fig:P-energy-sum}]
                {\includegraphics[width=\linewidth]{../images/P-energy-1} }
        }
    \caption{Time series for the period 1993 -- 2023 of (a) the yearly CE number of occurrences, (b) the yearly mean OI and (c) the yearly excess irradiation. The black lines represent the linear trends on the yearly data.}\label{fig:P-energy}
\end{figure}

@@

```{r relative-month-occurrences, echo=FALSE, fig.cap="Seasonal variability of the number of CE events in Thessaloniki for the period 1993 -- 2023 normalized to the maximum occurring in June, in the form of a box and whisker plot. The monthly values have also been normalized to the relative abundance of valid GHI observations. The box contains the data between the lower $25\\,\\%$ and the upper $75\\,\\%$ percentiles, the thick horizontal line and the diamond symbol represent the median and the mean values, respectively. The vertical lines (whiskers) extend between the maximum and minimum monthly values."}
knitr::include_graphics("../images/clim-CE-month-norm-MAX-median-N-MW-1.png")
```

@@

```{r ovir-distribution, echo=FALSE, fig.cap="Relative frequency distribution of CE events in Thessaloniki for the period 1993 -- 2023 as a function of OI. The histogram was split in two plots with different y-axis scales for better readability."}
knitr::include_graphics("../images/P-relative-distribution-diff-2.png")
```

@@

```{r ceg-duration-distribution, echo=FALSE, fig.cap="Relative frequency distribution of CE groups of consequent CE events according to their duration for Thessaloniki in the period 1993 -- 2023. The histogram was split in two plots with different y-axis scales for better readability."}
# knitr::include_graphics("../images/groups-1.png")
knitr::include_graphics("../images/groups-7.png")
```

@@

```{r group-2d, echo=FALSE, fig.cap="Relation of excess irradiation of CE groups with their duration for Thessaloniki in the period 1993 -- 2023. The logarithmic color scale denotes the frequency of the respective groups of events."}
knitr::include_graphics("../images/P-groups-bin2d-1.png")
```

@@

```{r relative-month-occurancies-ECE, echo=FALSE, fig.cap="Seasonal variability of the number of ECE events in Thessaloniki for the period 1993 -- 2023 normalized to the maximum occurring in March, in the form of a box and whisker plot. The box contains the data between the lower $25\\,\\%$ and the upper $75\\,\\%$ percentiles. The thick horizontal line and the diamond symbol represent the median and the mean values, respectively. The vertical lines (whiskers) extend between the maximum and minimum monthly values."}
knitr::include_graphics("../images/clim-ECE-month-norm-MAX-median-N-2.png")
```

@@

```{r P-extreme-distribution, echo=FALSE, fig.cap="Distribution of ECE events in Thessaloniki for the period 1993 -- 2023. Measured as extra irradiance over TSI at TOA for the same SZA."}
knitr::include_graphics("../images/P-extreme-distribution-2.png")
```

<!-- ```{r P-extreme-distribution, echo=FALSE, fig.cap="Distribution of ECE events in Thessaloniki for the period 1993 -- 2023."} -->
<!-- knitr::include_graphics("../images/P-extreme-distribution-1.png") -->
<!-- ``` -->

@@

By creating a clear sky approximation representing the long- and short-term
variability of the expected clear sky GHI, we were able to identify cases of CE
events in Thessaloniki for the period 1993 -- 2023.  After analyzing the CE cases, we
found an increase of
$`r plus_sign(round(yearlyCE[var == "GLB_diff.N", slope], 1))`\pm `r round(2 * yearlyCE[var == "GLB_diff.N", slope.sd], 1)`\,\text{cases}/\text{year}$,
with the mean annual irradiation of the CE events increasing with a rate of
$`r plus_sign(round(yearlyCE[var == "GLB_diff.sum", slope], 1))`\pm `r round(2 * yearlyCE[var == "GLB_diff.sum", slope.sd], 1)`\,\text{kJ}/\text{year}$.
The most active months of CE events are May and June.  We found that continuous CE
events can last up to $140$ minutes, while the duration of $80\,\%$ of them is bellow
$5$ minutes.

>>>>

author:
  - name:        Athanasios N. Natsis
    email:       natsisphysicist@gmail.com
    affiliation: LAP
    correspondingauthor: true
    # footnote: 1
  - name:        Alkiviadis Bais
    email:       abais@auth.gr
    affiliation: LAP
  - name:        Charikleia Meleti
    email:       meleti@auth.gr
    affiliation: LAP
    # footnote: 1
address:
  - code:         LAP
    organization: Laboratory of Atmospheric Physics, Physics Department, Aristotle University of Thessaloniki
    # addressline:  1 main street
    city:         Thessaloniki
    postcode:     54124
    country:      Greece
  # - code:         Another University
  #   organization: Department
  #   addressline:  A street 29
  #   postcode:     2054 NX
  #   city:         Manchester,
  #   state:        State
  #   country:      The Netherlands
# footnote:
#   - code: 1
#     text: "This is the first author footnote."
#   - code: 2
#     text: "Another author footnote."
abstract: |
  In this study we investigate the characteristics of global horizontal irradiance
  enhancement events induced by clouds over Thessaloniki for the period 1994 -- 2023
  using data recorded every one minute. We identified the cloud enhancement (CE)
  events by creating an appropriate clear-sky irradiance reference with the use of a
  radiative transfer model and aerosol optical depth data from a collocated Cimel sun
  photometer and a Brewer spectrophotometer. We found a trend in CE events of
  \(45.6\pm 21.9\,\text{cases}/\text{year}\), and a trend in the CE events
  irradiation of \(116.9\pm 67.8\,\text{kJ}/\text{year}\). The peak of the CE events
  was observed during May and June. The analysis of the total duration of CE events
  showed that durations longer than 5 minutes are very rare, with exceptions lasting
  over an hour and up to 140 minutes. Finally, we have detected enhancements above
  the total solar irradiance at the top of the atmosphere of up to
  \(400\,\text{W}/\text{m}^{2}\), with the \(75\,\%\) of the cases below
  \(200\,\text{W}/\text{m}^{2}\). The most active period of these extreme events is
  spring -- early summer with a secondary peak in autumn.

@@

We studied the enhancement of the global horizontal irradiance induced by clouds over
Thessaloniki for the period 1994 -- 2023 using data recorded every one minute.  We
identified the cloud enhancement events (CE) by creating an appropriate clear sky
irradiance reference with the use of a radiative transfer model and aerosol optical
depth data from a collocated Cimel sun photometer and a Brewer spectrophotometer. We
found a trend in CE events of
$`r plus_sign(round(yearlyCE[var=="GLB_diff.N",slope],0))`\pm
`r round(2*yearlyCE[var=="GLB_diff.N",slope.sd],0)`\,\text{cases}/\text{year}$,
and a trend in the CE events irradiation of 
$`r plus_sign(round(yearlyCE[var=="GLB_diff.sum",slope],1))`\pm
`r round(2*yearlyCE[var=="GLB_diff.sum",slope.sd],1)`\,\text{kJ}/\text{year}$.
The peak of the CE events was observed during May and June.  An analysis of the total
duration of CE events, showed that durations longer than 5 minutes are very rare,
with exceptions lasting over an hour and up to 140 minutes.  Finally, we have
detected enhancements above the total solar irradiance at the top of the atmosphere
for the same solar zenith angle of up to
$400\,\text{W}/\text{m}^{2}$, with the $75\,\%$ of the cases below
$200\,\text{W}/\text{m}^{2}$. The most active season of these extreme events is
spring until early summer and autumn.

@@

<!-- Radiation data quality check -->
To further improve the quality of the irradiance data, a manual screening was
performed, in order to remove inconsistent and erroneous recordings that can occur
stochastically or systematically during the long operation of the instrument.  The
manual screening was aided by a radiation data quality assurance procedure, adjusted
for the site, which was based on the methods of Long and Shi\ [@Long2006;
@Long2008a].  Thus, problematic recordings have been excluded from further
processing.  Furthermore, due to the significant measurement uncertainty in GHI when
the Sun is near the horizon, and due to some systematic obstructions by nearby
buildings, we have excluded all measurements with solar zenith angle (SZA) greater
than $`r 90 - BIO_ELEVA`^\circ$.  Finally, images from a sky camera have been used in
the manual inspection of the CE identification. The sky camera operates since 2012
and stores images in 5 min time steps.
<!--
Although it is impossible to detect all false data, the large number of
available data, and the aggregation scheme we used, ensures the quality of the
radiation measurements used in this study.
-->
<!-- Data selection -->
<!--
To preserve an unbiased representation of the data we applied a constraint, similar
the one used by @CastillejoCuberos2020. For each valid hour of day, there must exist
at least 45 minutes of valid measurements, including nighttime, when near sunrise and
sunset. Days with less than 5 valid hours were rejected completely.
-->
<!--
# CastillejoCuberos2020 There must be at least 5 h of valid data (Tests 1–7),
otherwise all data for this day is flagged as invalid. There must be at least This
work valid data 45 min of valid data (Tests 1–7) for an hour to be flagged as valid,
otherwise all minute data for this hour is flagged as invalid.
-->
An overview of the data analyzed for the study of the CE and ECE events is given in
Figure\nobreakspace{}\ref{fig:CLB-daily}, as the daily mean of the measured GHI.

```{r CLB-daily, echo=FALSE, fig.cap="Daily mean values of GHI in Thessaloniki for the period 1994 -- 2023."}
knitr::include_graphics("../images/P_daily_trend-3.png")
```

@@

<!-- Aerosol change -->
The clear-sky reference values discussed above are based on the climatological AOD
and WC; hence they cannot describe accurately the long-term variation of
$\text{GHI}_\text{ref}$ due to long-term changes in the two atmospheric constituents,
mainly AOD.  As reported by @Natsis2023, there is a long-term brightening effect in
the GHI data of Thessaloniki for the period 1993 -- 2023, which for clear-sky data
was attributed to long-term changes in aerosol effects.  Therefore, an adjustment of
the $\text{GHI}_\text{ref}$ during the period of study was made using simulations
with the RTM based on the long-term variations of the AOD. As AERONET data start only
in 2003, we used for the period 1994 -- 2005 estimates of changes in AOD at
$340\,\text{nm}$ derived from a collocated Brewer spectrophotometer [@Kazadzis2007]
to calculate the change in $\text{GHI}_\text{ref}$ due to aerosols during this period.

@@

```{r CS-change, echo=FALSE, fig.cap="Simulated long-term change in clear sky irradiance relative to the climatological values due to changes in AOD in Thessaloniki for the period 1994 -- 2023."}
knitr::include_graphics("../images/P-CS-change-1.png")
```

@@

Figure\nobreakspace{}\ref{fig:P-energy} shows the time series of the yearly number of
CE cases (each with duration of one minute), the yearly mean OI and the yearly
excess irradiation for the period 1994 -- 2023, together with corresponding linear
trends. All three quantities show increasing trends, most pronounced for the
frequency of occurrence 
($`r plus_sign(round(yearlyCE[var == "GLB_diff.N", slope], 0))`\pm `r round(2 * yearlyCE[var == "GLB_diff.N", slope.sd], 0)`\,\text{cases}/\text{year}$)
and the excess irradiation 
($`r plus_sign(round(yearlyCE[var == "GLB_diff.sum", slope], 1))` \pm `r round(2 * yearlyCE[var == "GLB_diff.sum", slope.sd], 1)`\,\text{kJ}/\text{year}$),
which are also statistically significant.
In contrast the trend of the yearly mean OI is negligible 
($`r plus_sign(round(yearlyCE[var == "GLB_diff.mean", slope], 2))`\pm `r round(2 * yearlyCE[var == "GLB_diff.mean", slope.sd], 1)`\,\text{W}/\text{m}^2/\text{year}$)
and of no statistical significance.
The average OI for the entire period is
$`r plus_sign(round(mean(ST_E_yearly$GLB_diff.mean), 1))`\pm `r round(sd(ST_E_yearly$GLB_diff.mean),1)`\,\text{W}/\text{m}^2$.
The interannual variability of the data about the trend lines is quite large.
Furthermore, the spread tends to increase with time (at least for the quantities of
panels b and c), suggesting a significant variability in cloud patterns over the
area, possibly associated to changes in climate.

@@

\begin{figure}% [h!]
        {\centering 
            \subfloat[\label{fig:P-energy-mean}]
                {\includegraphics[width=\linewidth]{../images/P-energy-3} }\\
            \subfloat[\label{fig:P-energy-N}]
                {\includegraphics[width=\linewidth]{../images/P-energy-2} }\\
            \subfloat[\label{fig:P-energy-sum}]
                {\includegraphics[width=\linewidth]{../images/P-energy-1} }
        }
    \caption{Time series for the period 1994 -- 2023 of (a) the yearly CE number of occurrences, (b) the yearly mean OI and (c) the yearly excess irradiation. The black lines represent the linear trends on the yearly data.}\label{fig:P-energy}
\end{figure}

@@

```{r relative-month-occurrences, echo=FALSE, fig.cap="Seasonal variability of the number of CE events in Thessaloniki for the period 1994 -- 2023 normalized to the maximum occurring in June, in the form of a box and whisker plot. The monthly values have also been normalized to the relative abundance of valid GHI observations. The box contains the data between the lower $25\\,\\%$ and the upper $75\\,\\%$ percentiles, the thick horizontal line and the diamond symbol represent the median and the mean values, respectively. The vertical lines (whiskers) extend between the maximum and minimum monthly values."}
knitr::include_graphics("../images/clim-CE-month-norm-MAX-median-N-MW-1.png")
```

@@

```{r ovir-distribution, echo=FALSE, fig.cap="Relative frequency distribution of CE events in Thessaloniki for the period 1994 -- 2023 as a function of OI. The histogram was split in two plots with different y-axis scales for better readability."}
knitr::include_graphics("../images/P-relative-distribution-diff-2.png")
```

@@

```{r ceg-duration-distribution, echo=FALSE, fig.cap="Relative frequency distribution of CE groups of consequent CE events according to their duration for Thessaloniki in the period 1994 -- 2023. The histogram was split in two plots with different y-axis scales for better readability."}
# knitr::include_graphics("../images/groups-1.png")
knitr::include_graphics("../images/groups-7.png")
```

@@

```{r group-2d, echo=FALSE, fig.cap="Relation of excess irradiation of CE groups with their duration for Thessaloniki in the period 1994 -- 2023. The logarithmic color scale denotes the frequency of the respective groups of events."}
knitr::include_graphics("../images/P-groups-bin2d-1.png")
```

@@

```{r relative-month-occurancies-ECE, echo=FALSE, fig.cap="Seasonal variability of the number of ECE events in Thessaloniki for the period 1994 -- 2023 normalized to the maximum occurring in March, in the form of a box and whisker plot. The box contains the data between the lower $25\\,\\%$ and the upper $75\\,\\%$ percentiles. The thick horizontal line and the diamond symbol represent the median and the mean values, respectively. The vertical lines (whiskers) extend between the maximum and minimum monthly values."}
knitr::include_graphics("../images/clim-ECE-month-norm-MAX-median-N-2.png")
```

@@

```{r P-extreme-distribution, echo=FALSE, fig.cap="Distribution of ECE events in Thessaloniki for the period 1994 -- 2023. Measured as extra irradiance over TSI at TOA for the same SZA."}
knitr::include_graphics("../images/P-extreme-distribution-2.png")
```

<!-- ```{r P-extreme-distribution, echo=FALSE, fig.cap="Distribution of ECE events in Thessaloniki for the period 1994 -- 2023."} -->
<!-- knitr::include_graphics("../images/P-extreme-distribution-1.png") -->
<!-- ``` -->

@@

By creating a clear sky approximation representing the long- and short-term
variability of the expected clear sky GHI, we were able to identify cases of CE
events in Thessaloniki for the period 1994 -- 2023.  After analyzing the CE cases, we
found an increase of
$`r plus_sign(round(yearlyCE[var == "GLB_diff.N", slope], 1))`\pm `r round(2 * yearlyCE[var == "GLB_diff.N", slope.sd], 1)`\,\text{cases}/\text{year}$,
with the mean annual irradiation of the CE events increasing with a rate of
$`r plus_sign(round(yearlyCE[var == "GLB_diff.sum", slope], 1))`\pm `r round(2 * yearlyCE[var == "GLB_diff.sum", slope.sd], 1)`\,\text{kJ}/\text{year}$.
The most active months of CE events are May and June.  We found that continuous CE
events can last up to $140$ minutes, while the duration of $80\,\%$ of them is bellow
$5$ minutes.
